Anatomical basis of sex differences in the electrocardiogram identified by three-dimensional torso-heart imaging reconstruction pipeline

Version 1

Abstract

Abstract

The electrocardiogram (ECG) is used for diagnosis and risk stratification following myocardial infarction (MI). Women have a higher incidence of missed MI diagnosis and complications following infarction, and to address this we aim to provide quantitative information on sex-differences in ECG and torso-ventricular anatomy features. A novel computational automated pipeline is presented enabling the three-dimensional reconstruction of torso-ventricular anatomies for 425 post-MI subjects and 1051 healthy controls from UK Biobank clinical images. Regression models were created relating torso-ventricular and ECG parameters. For post-MI women, the heart is positioned more posteriorly and vertically, than in men (with healthy women yet more vertical). Post-MI women exhibit less QRS prolongation, requiring 27% more prolongation than men to exceed 120ms. Only half of the sex difference in QRS is associated with smaller female cavities. Lower STj amplitude in women is striking, associated with smaller ventricles, but also more superior and posterior cardiac position. Post-MI, T wave amplitude and R axis deviations are strongly associated with a more posterior and horizontal cardiac position in women (but not in men). Our study highlights the need to quantify sex differences in anatomical features, their implications in ECG interpretation, and the application of clinical ECG thresholds in post-MI.

Introduction

The electrocardiogram (ECG) is a vital tool for routine clinical assessment of cardiac electrical abnormalities. Patient differences in anatomy and electrophysiology both determine the ECG. Computational and experimental studies have demonstrated that torso-ventricular orientation can substantially affect the ECG, potentially confounding disease diagnosis, and anatomy rather than electrophysiology is the strongest contributor to ECG variability in healthy subjects [1–4]. Sex and body mass index (BMI) are strongly associated with differences in torso and heart anatomy, and thus it is crucial to take them into consideration in ECG interpretation [5–7].

Ischemic heart disease is the leading single cause of mortality worldwide, killing 0.1% of the global population per year [8]. Therefore, effective patient diagnosis of myocardial infarction (MI) and risk stratification for treatment selection are crucial for survival and quality of life [9]. There is a higher incidence of missed diagnosis of acute MI in women, and hospital care can be less aggressive [10–12]. This is a potential contributor to the higher female mortality following MI [13], alongside factors such as differing demographic risk characteristics and comorbidities [14, 15].

Current guidelines using primarily left ventricular ejection fraction (LVEF) fail to identify the majority of sudden cardiac death (SCD) cases, prompting calls for further research [16]. Several ECGbased biomarkers have been proposed based on mechanistic knowledge on the underlying electrophysiological substrate post-MI. QRS duration and its dispersion have been a key component of proposed electrocardiographic risk stratification for SCD following MI. However, findings of its effectiveness are mixed, with some trials suggesting it is a predictor of arrhythmic risk following MI [17, 18], but others finding that it is more likely a predictor of general mortality than arrhythmic inducibility [19, 20].

T wave amplitude and ST segment changes are also important, reflecting repolarisation heterogeneity [21] and ST-elevation MI (STEMI) mandating immediate reperfusion therapy [22]. ECG abnormalities are caused by alterations in the electrophysiological substrate introduced by the infarct, but importantly they are modulated by anatomical factors such as heart size, location and orientation with respect to the ECG leads [2, 3].

Women’s ECGs typically exhibit shorter QRS durations and lower ST junction (STj) and T wave amplitudes, but longer QT intervals, than male [23]. Both electrophysiological and structural factors contribute to these differences. Testosterone increases repolarising potassium currents and decreases L-type calcium currents, both acting to speed up repolarisation [5]. Additionally, left ventricular mass is smaller in women than men [24], explaining in part the shorter QRS duration and lower STj and T wave amplitudes [25, 26]. Smaller torso volume can also lead to larger QRS and T wave amplitudes in women [1, 2] , but the impact on STj amplitude and whether anatomy is the only factor explaining these differences is unknown. In women, electrophysiological remodelling postinfarction may be less substantial than in men, partially from the protective effects of estrogen [27]. Investigations are urgently needed to fill these gaps for improvements in patient diagnosis, risk stratification, and treatment selection.

This study aims to quantify sex-differences in heart-torso anatomy and ECG biomarkers, as well as their relative associations, in healthy and post-MI subjects. To enable a population-wide study of heart-torso anatomy, we first develop and validate an automated image processing pipeline to reconstruct three-dimensional torso anatomical models using clinically standard cardiac magnetic resonance (CMR) images. We hypothesise that sex-differences in ECG biomarkers are only partially explained by smaller female heart and torso anatomies, and that sex-differences in electrophysiology and torso-hear orientation are particularly important, especially post-MI. These insights will lay the groundwork for a process to correct for the impact of torso-ventricular anatomy on the ECG, enabling personalised, automated ECG interpretation.

Methods

Computational pipeline

Figure 1A illustrates the fully automated pipeline used to create personalised torso anatomical reconstructions from clinically standard CMR imaging. This integrates and reformulates state-of-the-art image processing techniques in order to give the first fully automated, and extensively validated, torso-ventricular reconstruction pipeline of its kind.

Structure and validation of novel, automated torso reconstruction pipeline.

A: proposed end-to-end automated 3D torso reconstruction pipeline from 2D standard clinical cardiac magnetic resonance scans. The torso contours are first extracted from the images. A SSM is fitted to the contours, which is then optimally deformed. SSM: statistical shape model, α: spin of the cardiac short axis plane around the torso vertical axis, β: verticality of the cardiac long axis, γ: tilt of the cardiac short axis plane. B: (i) comparison of Dice coefficient between single stage contouring (blue) and 3 stage segmentation, postprocessing and refinement (red). (ii) surface-to-contour distance for an example case between the automatically reconstructed surface and the automatically generated torso contours. (iii) surface-to-surface distance between the torso mesh created using the automated pipeline and the manually annotated contours for the subjects with the smallest (left) and largest (right) electrode error respectively.

In short, the pipeline combines machine learning-based segmentation and contour extraction, statistical shape modelling and automated mesh deformation in the following steps:

Three-stage torso contour extraction from CMR imaging using machine learning to form sparse torso representation

Approximately 60 localiser and scout images per subject were first segmented using a convolutional neural network; then their outline was extracted with automated post-processing, and finally refined using a second network. U-Net [28] was used for both network architectures, as described in Smith et al. [29].

Fitting sparse representation contours to a statistical shape model of the human body to give an initial fitted torso

To generate torso meshes from sparse torso contours, a statistical shape model (SSM) of human body shapes made from a large optical scan dataset was fitted [30, 31]. As described in Supplementary Appendix 1.2.1, the principal components of the SSM were iteratively adjusted, alongside rigid transformations of the SSM in order to minimise distance between the surface and contours.

Automated deformation of the initial torso towards the extracted contours using thin plate splines to give a final dense torso mesh

As the fitted surface is inherently limited by variability in the SSM, a final deformation step was applied in order to better capture inter-subject variation. A thin plate splines method was used to pull the surface towards the contours, whilst retaining smoothness and regularity [32], as described in Supplementary Appendix 1.2.2. Laplacian smoothing, decimation, and affine transformation were applied to ensure the local geometric and topological characteristics of the reconstructed mesh.

Integration with a sparse cardiac mesh and subsequent extraction of torso-ventricular anatomical parameters

Cardiac surfaces were segmented using the method in Banerjee et al. [33] and torso-ventricular anatomical parameters were extracted, as detailed in Supplementary Appendix 1.3. The torso volume and two metrics of cardiac size were estimated. The locations of the ECG electrodes were identified on the mean SSM torso, and their positions were transformed with the mesh such that the resulting torso had electrodes located in equivalent locations. The position of the heart centre relative to the electrodes was estimated in the x (lateral), y (posterior) and z (superior) directions. The orientation of the heart was quantified by the Euler angles of the cardiac axes with respect to the torso axes. This includes the spin of the cardiac short axis plane around the torso vertical axis, the verticality of the cardiac long axis, and the tilt of the cardiac short axis plane.

Validation of the torso reconstruction pipeline

The CMR images of 30 subjects, chosen to be representative of the underlying dataset, were manually annotated to form an independent test set for validation. Final results of the three-stage torso contour extraction process were compared with simply using a single U-Net architecture to process raw images directly to refined contours in order to assess the accuracy of the torso-air boundary placement. Separate reconstructions of the 30 test subjects were made using the fully automated pipeline and using manually annotated contours. For each subject, the surface-to-surface distance between the two reconstructions was calculated to evaluate error propagation from contouring to reconstruction, as detailed in Supplementary Appendix 1.4. The mean surface-to-contour distance was calculated for each test subject to evaluate the quality of the reconstruction.

Dataset

The UK Biobank (UKB) dataset [34] was chosen for its high quality CMR imaging and 12-lead ECGs in a large population. 1646 control subjects were randomly selected from the UKB cohort alongside the 479 subjects with a history of MI, as defined in Supplementary Appendix 1.1. The healthy cohort was made by excluding control subjects with disease diagnoses, and subjects in either population were excluded if their ECG or imaging data was not complete, as described in Supplementary Appendix 1.1. The final population for reconstruction and statistical analysis comprised 1051 healthy and 425 post-MI subjects (total 1476), as shown in Tables 1 and 2. 92.7% of subjects identified as White British. The post-MI population was 6.1±0.4 years older and had 2.4±0.2 kg/m2 higher BMI, with no statistically significant differences in ethnicity. This data was obtained through UK Biobank Resource under Application Number ‘40161’ and the data was processed according to their guidelines. All participants gave written informed consent before enrolment in the UK Biobank cohort.

Continuous variables are shown as mean ± standard deviation and categorical variables as number of subjects (percentage in bracket). The p-value refers to the null hypothesis that the mean, or proportion in a category, is equal in the healthy and MI subpopulations. Distributions of continuous variables were tested for normality and equal variance, and the appropriate statistical test was selected, as outlined in the statistical methods Supplementary Appendix 1.5. χ2-tests were performed to assess the proportions of subjects for categorical variables. MI: myocardial infarction, BMI: body mass index.

Categorical variables as number of subjects (percentage in bracket). χ2-tests were performed to assess the proportions of subjects with each disease or treatment. The p-value refers to the null hypothesis that the proportion in a category is equal in the male and female subpopulations. Surgical treatment information is sourced from hospital records, whereas disease information is obtained from multiple sources including hospital records, primary care and self-report.

Statistical analysis

ECG parameters were obtained from the UK Biobank database, which were automatically extracted with the proprietary software Cardiosoft [35]. The distribution of each anatomical and ECG parameter was compared between populations as in Supplementary Appendix 1.5. STj amplitude refers to the amplitude of the ST segment measured at the junction point (the end of the QRS complex). The contribution of each anatomical parameter to the sex difference in ECG biomarkers, and the remaining estimated effect of electrophysiology, were calculated by regression analysis, as detailed in Supplementary Appendix 1.5. Similarly, their contribution to the difference between healthy and post-MI ECG biomarkers was estimated. A linear model was trained to estimate, and therefore correct, the total anatomical contribution to ECG parameters using only demographic features, as described in Supplementary Appendix 1.5.

Results

Reconstructed torso surfaces matched input contours to within 2.19mm

Figure 1B showcases the capability of the computational pipeline to extract torso contours and reconstruct them into 3D meshes. The three-stage segmentation, automated post-processing, and refinement procedure improved the Dice coefficient of torso contour extraction, compared to a single-stage contouring network, across all image views (Figure 1B(i)). The median surface-to-contour distance, showing the quality of the reconstructed torso surface (Figure 1B(ii)), for the test subjects was 2.19mm (inter-quartile range - IQR: 1.98–2.37mm). The median surface-to-surface distance between torso meshes reconstructed using the manually annotated contours and the machine learning pipeline for the 30 test subjects was 0.82mm (IQR: 0.69-1.10mm). Figure 1B(iii) shows the reconstructed surfaces for the test subjects with the smallest and largest median reconstruction errors. Both reconstructed surfaces were smooth and anatomically realistic, and, across most of the torsos, errors are within the millimetre range.

Torso volume was larger in post-MI than in healthy subjects in women and men (by 3.7±0.9 dm3 (10.2±3.3%) and 3.5±0.5 dm3 (8.2±1.9%) respectively (Supplementary Figure 3). Additionally, for women, in post-MI the heart was positioned more posteriorly by 3.1±0.9mm, than in healthy subjects. The heart was also oriented with its long axis more horizontally by 6.0±1.2° and 4.3±0.5° in post-MI versus healthy women and men, respectively. These changes were related to the increased BMI of post-MI subjects (Supplementary Figure 3K). Age was also considered (Supplementary Figure 3J), but its effects were considerably less substantial.

QRS duration was shorter in women than in men, particularly post-MI, due to smaller hearts and reduced impact of electrophysiological abnormalities

For both healthy and post-MI subjects, QRS duration was shorter in women versus men in all leads (Figures 2A and 2B), with contributions from each anatomical factor displaying sex differences (Figures 2C and 2D). Each factor’s contribution is calculated from the product of the regression coefficients and anatomical sex differences (Supplementary Appendix 1.5). Sex differences in cavity volume explained approximately half of the sex difference in QRS duration in both healthy and postMI subjects (a mean value across all leads of 3.4±1.3ms out of the 6.0±1.5ms mean total in healthy, and 4.5±1.4ms of a total sex difference of 8.3±2.5ms in post-MI). In healthy subjects the electrophysiological contribution to QRS sex differences was similar in all precordial leads (1.0±1.0ms to 2.3±1.2ms). In post-MI subjects, it was considerably more variable (0.2±2.8ms in V1 to 6.3±3.0ms in V6), possibly due to the infarct area.

Sex differences in QRS duration in healthy and post-MI subjects.

A, B: mean QRS duration for each ECG lead in healthy and post-MI subjects respectively with women (red circles) and men (cyan triangles). C,D: contribution of anatomical parameters and electrophysiology to sex differences in QRS duration (QRSd) for healthy and post-MI subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between male and female populations). E, F: mean QRS durations for each ECG lead in men and women respectively for healthy (black squares) and post-MI (red diamonds). G,H: contribution of anatomical parameters and electrophysiology to differences in QRS duration (QRSd) between healthy and post-MI subjects for male and female subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between healthy and post-MI populations). Female y-axis limits have been adjusted by the difference in healthy QRS duration between sexes for ease of comparison.

Post-MI women exhibit less QRS prolongation than men (Figures 2E and 2F), with mean percentage increase in QRS duration from healthy to post-MI, across all leads, of 3.4±2.3% and 5.8±1.5% respectively. This is explained by a reduced contribution of electrophysiological abnormalities post-MI in women, as shown in Figures 2G and 2H (smaller grey bars).

For lead V6, 15.9% of healthy subjects had an estimated anatomical QRS difference of over 5ms from a subject with mean anatomical parameters. After correction by a linear model that was trained only on demographic features (sex, height, BMI, and age), this proportion reduced to 5.1%.

STj amplitude is lower in women than men in both healthy and post-MI subjects, explained by smaller hearts in more superior and posterior position

STj amplitude was lower in women versus men, particularly in the septal and anterior leads in healthy and post-MI subjects (Figures 3A and 3B). As for QRS duration, this can be partly explained by women’s smaller LV mass (by 29.3±5.4% and 27.1±5.7% in healthy and post-MI subjects respectively), but also by a more superior position for healthy subjects (by 9.6±0.9mm) and more posterior position for post-MI subjects (by 4.3±1.0mm) (Figures 3C and 3D). Both smaller LV mass and electrophysiology contribute to lower STj amplitude, whereas a smaller torso volume was generally associated with a higher STj amplitude (yellow and grey versus orange bars in Figures 3C and 3D). This pattern was consistent for differing measurement times of the ST level (Supplementary Figure 6). Baseline STj amplitude in lead V4 was 0.028±0.003mV lower in healthy women than healthy men. Healthy women with BMI>25kgm-2 had a STj amplitude 0.047±0.004mV lower than healthy men with BMI<25kgm-2associated with their smaller ventricles, but larger torsos.

Sex differences in ECG amplitudes in healthy and post-MI subjects.

A, B: mean STj amplitude (measured at QRS offset) for each ECG lead in healthy and post-MI subjects respectively with women (red circles) and men (cyan triangles). C, D: contribution of anatomical parameters and electrophysiology to sex differences in STj amplitude for healthy and post-MI subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between male and female populations). E, F: mean T wave amplitude (TWA) for each ECG lead in men and women respectively for healthy (black squares) and post-MI (red diamonds). G, H: contribution of anatomical parameters and electrophysiology to differences in T wave amplitude (TWA) between healthy and post-MI subjects for male and female subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between healthy and post-MI populations). Female y-axis limits have been adjusted by the difference in healthy T wave amplitude between sexes for ease of comparison.

T wave amplitude (TWA) following MI is more strongly associated with a more posterior cardiac position and more horizontal orientation in women than men

Precordial TWA was lower in women than men, by 41.7±20.3% and 31.2±20.0% in healthy and post-MI populations respectively (Supplementary Figure 7). MI resulted in reduced TWA, in all leads in men, whereas only in V4-6 and limb leads in women (Figures 3E and 3F). For men, this was largely associated with electrophysiological factors substantially reducing TWA across leads (Figure 3G). However, for women, the reduced precordial TWA was largely associated with more posterior and horizontal ventricles from their larger BMI (Figure 3H), and electrophysiological factors actually counteracted the TWA post-MI reduction for septal leads (shown as positive contributions).

R axis deviations following MI are associated with more horizontal cardiac orientation in women, but MI-related electrophysiological abnormalities in men

Whereas the previous ECG parameters can be assessed in individual leads, ECG-derived axis angles are a composite metric, positively correlated with amplitudes in leads such as lead aVF, and negatively correlated with lead I [36]. Figure 4A breaks down the anatomical contributions to R and T axis shifts in post-MI compared with healthy subjects. For men, the electrophysiological contribution dominates their 12.0±2.8° left deviated R axis following MI. However, for women their 5.3±4.0° left deviation in the R axis is mainly associated with a more horizontal cardiac long axis, with electrophysiological factors playing very little role (Figure 4B). This is partly explained by female R axes being more sensitive to changes in verticality - the regression coefficient between R axis and verticality is higher for women than men (Figure 4C). The female T axis was more right deviated than male, largely associated with electrophysiology (Figures 4A and 4B), with less contribution from orientation (Figure 4D).

Sex differences in ECG axis angles in healthy and post-MI subjects

A, B: contribution of anatomical parameters and electrophysiology to differences in R and T axis angles between healthy and post-MI subjects for male and female subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between healthy and post-MI populations). C, D: normalised regression coefficients showing the association between the verticality of the cardiac long axis and the R and T axes respectively for the healthy and post-MI populations. Bars shown in red represent regression coefficients significantly different from 0, at a significance level of 0.05. E, F: normalised regression coefficients showing the association between torso volume and the R and T axes respectively for the healthy and post-MI populations.

Torso volume significantly affected R axis for both sexes in healthy, but not post-MI subjects (Figure 4E, regression coefficient of R axis against torso volume). For example, in healthy women, every 10dm3 increase in torso volume was associated with a decrease in R axis of 7.2±3.0°. This was related to increased distance to the LL electrode (inversely related to the R axis) without as substantial an increase in distance to the LA electrode (positively related to the R axis). For healthy women, every 10dm3 increase in torso volume was associated with an increased distance to the heart centre of 19.4±0.7mm and 5.9±0.6mm for the LL and LA electrodes, respectively. However, in post-MI women, the gap between these differences was smaller, with corresponding increased distances of 16.6±1.8mm and 8.3±1.7mm. This association was less strong for the T wave (Figure 4F).

Discussion

This study provides quantification of sex differences in image-based anatomical factors and their implications for ECG biomarkers in healthy and post-MI subjects, enabled by a novel computational clinical image processing pipeline. With this work, we aim to contribute to addressing the higher incidence of missed diagnosis and increased female MI mortality following infarction [12, 13, 37]. Our analysis of a large cohort of healthy and post-MI subjects’ images and ECG data provides quantitative evidence on the importance of considering demographic characteristics and their anatomical differences, particularly sex, for corrections of ECG biomarkers, key to patient risk stratification

An important contribution is the novel computational pipeline for automated 3D torso reconstruction, which enables high-throughput investigations into the effects of anatomical variability on the study of a wide range of cardiac diseases. The adaptation of the pipeline to standard clinical cardiac imaging allows for the exploitation of extensive databases, such as the UK Biobank, and sets the groundwork for clinical tools to personalise ECG interpretation considering patient anatomy.

Our analysis provides the following insights into sex-related anatomical differences on the impact of myocardial infarction on key ECG biomarkers: 1) Half of the shorter QRS durations in women are explained by smaller cavity volumes in both healthy and post-MI subjects. 2) Female STj amplitude is lower than male in all precordial leads for healthy and post-MI subjects, associated with a smaller LV mass and more superior and posterior position. 3) In women, anatomical factors have a stronger effect in TWA changes following MI than in men. 4) R axis is more left deviated in MI than healthy populations; for women this is associated with a more horizontal cardiac long axis, but for men with electrophysiology.

Both clinically and moderately prolonged QRS durations have been associated with increased risk following MI [17, 18]. Importantly, our analysis reports that mean QRS duration would have to increase by 27% more for women than men to exceed 120ms and by 68% more to exceed 100ms, in lead aVL. Sex differences in QRS duration are only partly explained by cardiac size, in line with [1], and with the biophysical understanding that larger distances require longer to traverse [38]. Our findings suggest that corrected QRS duration taking into consideration demographics and their corresponding anatomical differences may improve risk stratification post-MI. QRS prolongation in post-MI subjects has been reported in previous studies, particularly cooccurring with ST changes in leads exploring the affected region [39]. It can be explained by slowed conduction velocity in the affected region, involvement of the conduction system such as the bundle branches and Purkinje, and in the chronic case, with dilation and hypertrophy of non-infarcted myocardium [19, 40, 41].

The finding that STj amplitude was substantially lower in women than men in all septal and anterior leads (V1-V4) is also striking. Based on our findings, the dual effects of cardiac and torso volume need to be considered in the interpretation of ST-elevation through the lens of the anatomical context of the patient. As we considered healthy and post-MI subjects (rather than with acute ischemia), our data reflect the normal baseline of STj amplitude, from which any deviations can be measured. The Fourth Universal Definition of Myocardial Infarction only has sex-specific thresholds in leads V2 and V3 for identifying ST-elevation, a key component in the diagnosis of ST-elevated myocardial infarctions (STEMIs) [42]. For example, in lead V4, the clinical threshold for both female and male ST-elevation is 0.1mV [42]. Based on our findings, due to their lower baseline values, overweight healthy women would need a 63% larger increase in STj amplitude to be classified as ST-elevated than normal weight healthy men. Therefore, this work represents a significant step forward in the characterisation of anatomical influences of its baseline level, from which deviations can then be more accurately identified in the context of anatomical variation. It also highlights the important intersectional effects of demographic characteristics such as sex and BMI, suggesting that both of these factors should be accounted for in assessing ST-elevation to improve accuracy of MI diagnosis and classification.

Alterations to the polarity and amplitude of the T wave are used in diagnosis of acute MI [42] and TWA affects proposed risk stratification tools, particularly markers of repolarisation abnormalities [9, 43]. Our data reveal that, particularly in women, a substantial proportion of TWA reductions are associated with anatomical differences, such as the decreased verticality of the postMI heart. This implies that without anatomical corrections, such as adjusting for BMI, the female TWA may be less representative of its underlying electrophysiology, and therefore less representative of any pathology. This increased effect of anatomical parameters in women is partly due to their larger difference in key anatomical factors such as posterior position and long axis orientation from healthy to post-MI subjects. This reflects their larger increase in BMI, associated with a more horizontal cardiac long axis, as described in previous works, suggesting this was related to an upwards shift in the diaphragm [6, 44]. Furthermore, the female TWA is more sensitive to changes in anatomical parameters. This could be explained by sex differences in the directions of the T wave vectors and nonlinearities in the structure-function relationships. Our finding that TWA is affected by cardiac position and orientation is in line with previous simulation studies [2] albeit in a small number of anatomies (N=5) and without electrophysiological variation.

Our data reveal that the verticality of the cardiac long axis is particularly significant in determining R axis angle, especially for women. This represents a step forward in understanding the complex relationship between electrical and anatomical axes [6, 45, 46]. Moreover, the leftdeviation of the R axis may look similar in both sexes; in men this is representative of an electrophysiological change, presumably alterations to the direction of the QRS vector caused by both hypertrophy and areas of electrical inactivity. However, for women this change is almost entirely associated with anatomical differences. This suggests that it is particularly important to correct R axis angles for anatomical factors in women to avoid misattribution of deviations to electrophysiological pathology. The significant relationship between torso volume and R axis angle for healthy subjects indicates that BMI may be a critical consideration when using ECG based biomarkers that are sensitive to axis angles. The fact that increased torso volume moved the LL electrode further away from the heart centre than the LA electrode suggests that this relationship is caused by a stretching of the anatomical plane in which these angles are viewed. Therefore, this represents another area in which a clinical tool that accounts for anatomical variation in interpreting ECG biomarkers would reduce differential accuracy in risk stratification between demographic groups.

The computational pipeline presented here is available to be applied to further studies. Validation is performed by demonstrating that the extracted contours are near the ground truth torso-air boundary on the clinical imaging, as well as to the final reconstructed surfaces (i.e. high Dice coefficient for contour extraction and low surface-to-contour distances). As detailed in Smith et al. [29], the three-step segmentation, automated post-processing, and refinement process was particularly effective at removing errors inside the torso at locations of sharp gradient between dark and light, for example due to air in the lungs. Whilst reconstructed 3D torso surfaces were largely smooth and realistic, there were small irregularities, generally on the shoulder and waist regions. These were not typically associated with incomplete CMR artifact removal and may instead result from images being taken on different breath-holds or the image sparsity. The low surface-to-surface distance between reconstructions made using the automatically and manually derived contours demonstrates that the small errors in contour extraction did not lead to significant errors in electrode placement on the torso meshes. Surfaces were generally more accurate on the upper left quadrant of the torso, which is where most electrodes are placed. The validity of the torso-ventricular reconstruction pipeline is further supported as known trends in torso and cardiac size, position, and orientation were qualitatively reproduced, and furthermore were quantitively characterised. This includes the more superior female cardiac position [47, 48] and the horizontal shift in the cardiac long axis for subjects with a larger BMI [6, 44]. As this pipeline uses clinically standard CMR imaging, it can be more easily translated to a clinical tool to correct for the effect of anatomical variation on the ECG in a variety of medical contexts.

Due to the demographics of the UK Biobank [34], this study only included subjects aged 4580 years old; however, the incidence of MI is very low in people younger than this range [13]. The ethnic background of the dataset also had limited diversity; however, this is roughly representative of the UK Biobank population and thus the ethnic composition of the UK at the time of recruitment [49]. As databases become more diverse, the mediating effect of ethnic background on the cardiac structure-function relationship should be investigated. Whilst the healthy subjects had a roughly equal sex balance, the post-MI subjects had a male-female ratio of approximately 4:1, in line with sex differences in diagnosed MI events [50]. The statistical shape model used for initial torso reconstruction did not have separate male and female models, which may have led to an underestimation in the sex differences in the torsos of male and female anatomies. However, the deformation step of the reconstruction was designed to address this issue by modelling personalised torso variations. The exact electrode positions of the obtained ECGs were not available in the UK Biobank dataset, so these were estimated by placing the electrodes on standard positions on the personalised torso. However, this estimation process allows for use of many similar clinically standard datasets, avoiding complex acquisition protocols. As the anatomical factors cannot represent a perfectly complete and independent basis set, the electrophysiological contribution can only be estimated. However, as there is more anatomical variation than can be represented using the given factors, such as the smaller scale geometry of the heart, the role of the anatomy versus electrophysiology was likely under-rather than overestimated. Whilst the study benefits from the electrophysiological and anatomical variation that using clinical data facilitates, it is impossible to fully separate the effects of these factors on ECG biomarkers, as characteristics such as sex affect both categories, making them interdependent. This means that this work complements computational studies, which have less interpersonal variability, but isolate the effect of positional and rotational changes. Scar characteristics were not available in the UK Biobank, but the effect of scar geometry variation could be considered in future studies with the development of scar delineation tools. Future work should focus on the characterisation of ECG biomarkers in myocardial infarction and other disease conditions. The automated torso reconstruction pipeline this work proposes enables the expansion of the number of anatomies such future computational studies would investigate.

Conclusion

To conclude, this work presents a novel automated pipeline for personalised torso reconstruction and demonstrates its power in exploiting large clinical databases by relating torso-ventricular anatomy and ECG biomarkers for 1051 healthy and 425 post-MI subjects. Results show the considerable influence of anatomical factors in demographic differences of ECG biomarkers. Women and individuals with higher BMI may be disproportionately affected. This underlines the importance of, and lays the foundations for, personalised ECG interpretation that considers a subject’s individual torso-ventricular anatomy, facilitating improvement in diagnosis and risk stratification tools.

eLife Assessment

This important study combines electrocardiographic (ECG) and heart/torso anatomy data from subjects included in the UK Biobank to analyze sex-specific differences in relationships between those two characteristics. The study has several compelling strengths, including the development of an open-source pipeline for reconstruction and analysis of heart/torso geometry from a large cohort. Nevertheless, technical analysis of the data as presented is incomplete, specifically as it pertains to assessment of co-linearity between regressed parameters, interpretation of regression coefficients for sex and/or presence of myocardial infarction, and discussion of potential roles played by underlying electrophysiological derangements. With improvements to these aspects of the analysis, the paper would be of interest to the cardiovascular research community, especially those studying highly relevant health and treatment disparities arising from sex differences.

Reviewer #1 (Public review):

Summary:

The electrocardiogram (ECG) is routinely used to diagnose and assess cardiovascular risk. However, its interpretation can be complicated by sex-based and anatomical variations in heart and torso structure. To quantify these relationships, Dr. Smith and colleagues developed computational tools to automatically reconstruct 3D heart and torso anatomies from UK Biobank data. Their regression analysis identified key sex differences in anatomical parameters and their associations with ECG features, particularly post-myocardial infarction (MI). This work provides valuable quantitative insights into how sex and anatomy influence ECG metrics, potentially improving future ECG interpretation protocols by accounting for these factors.

Strengths:

(1) The study introduces an automated pipeline to reconstruct heart and torso anatomies from a large cohort (1,476 subjects, including healthy and post-MI individuals).

(2) The 3-stage reconstruction achieved high accuracy (validated via Dice coefficient and error distances).

(3) Extracted anatomical features enabled novel analyses of disease-dependent relationships between sex, anatomy, and ECG metrics.

(4) Open-source code for the pipeline and analyses enhances reproducibility.

Weaknesses:

(1) The linear regression approach, while useful, may not fully address collinearity among parameters (e.g., cardiac size, torso volume, heart position). Although left ventricular mass or cavity volume was selected to mitigate collinearity, other parameters (e.g., heart center coordinates) could still introduce bias.

(2) The study attributes residual ECG differences to sex/MI status after controlling for anatomical variables. However, regression model errors could distort these estimates. A rigorous evaluation of potential deviations (e.g., variance inflation factors or alternative methods like ridge regression) would strengthen the conclusions.

(3) The manuscript's highly quantitative presentation may hinder readability. Simplifying technical descriptions and improving figure clarity (e.g., separating superimposed bar plots in Figures 2-4) would aid comprehension.

(4) Given established sex differences in QTc intervals, applying the same analytical framework to explore QTc's dependence on sex and anatomy could have provided additional clinically relevant insights.

Reviewer #2 (Public review):

Summary:

Missed diagnosis of myocardial ischemia (MI) is more common in women, and treatment is typically less aggressive. This diagnosis stems from the fact that women's ECGs commonly exhibit 12 lead ECG biomarkers that are less likely to fall within the traditional diagnostic criteria. Namely, women have shorter QRS durations and lower ST junction and T wave amplitudes, but longer QT intervals, than men. To study the impact, this study aims to quantify sex differences in heart-torso anatomy and ECG biomarkers, as well as their relative associations, in both pre- and post-MI populations. A novel computational pipeline was constructed to generate torso-ventricular geometries from cardiac magnetic resonance imaging. The pipeline was used to build models for 425 post-myocardial infarction subjects and 1051 healthy controls from UK Biobank clinical images to generate the population.

Strengths:

This study has a strength in that it utilizes a large patient population from the UK Biobank (425 post-MI and 1051 healthy controls) to analyze sex-based differences. The computational pipeline is state-of-the-art for constructing torso-ventricular geometries from cardiac MR and is clinically viable. It draws on novel machine learning techniques for segmentation, contour extraction, and shape modeling. This pipeline is publicly available and can help in the large-scale generation of anatomies for other studies. This allows computation of various anatomical factors (torso volume, cavity volume, etc), and subsequent regression analysis on how these factors are altered before and after MI from the 12-lead ECG.

Weaknesses:

Major weaknesses stem from the fact that, while electrophysiological factors appear to play a role across many leads, both post-MI and healthy, the electrophysiological factors are not stated or discussed. The computational modeling pipeline is validated for reconstructing torso contours; however, potential registration errors stemming from ventricular-torso construction are not addressed within the context of anatomical factors, such as the tilt and rotation of the heart. This should be discussed as the paper's claims are based on these results. Further analysis and explanation are needed to understand how these sex-specific results impact the ECG-based diagnosis of MI in men and women, as stated as the primary reason for the study at the beginning of the paper. This would provide a broader impact within the clinical community. Claims about demographics do not appear to be supported within the main manuscript but are provided in the supplements. Reformatting the paper's structure is required to efficiently and effectively present and support the findings and outcomes of this work.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

The electrocardiogram (ECG) is routinely used to diagnose and assess cardiovascular risk. However, its interpretation can be complicated by sex-based and anatomical variations in heart and torso structure. To quantify these relationships, Dr. Smith and colleagues developed computational tools to automatically reconstruct 3D heart and torso anatomies from UK Biobank data. Their regression analysis identified key sex differences in anatomical parameters and their associations with ECG features, particularly post-myocardial infarction (MI). This work provides valuable quantitative insights into how sex and anatomy influence ECG metrics, potentially improving future ECG interpretation protocols by accounting for these factors.

Strengths:

(1) The study introduces an automated pipeline to reconstruct heart and torso anatomies from a large cohort (1,476 subjects, including healthy and post-MI individuals).

(2) The 3-stage reconstruction achieved high accuracy (validated via Dice coefficient and error distances).

(3) Extracted anatomical features enabled novel analyses of disease-dependent relationships between sex, anatomy, and ECG metrics.

(4) Open-source code for the pipeline and analyses enhances reproducibility.

Weaknesses:

(1) The linear regression approach, while useful, may not fully address collinearity among parameters (e.g., cardiac size, torso volume, heart position). Although left ventricular mass or cavity volume was selected to mitigate collinearity, other parameters (e.g., heart center coordinates) could still introduce bias.

(2) The study attributes residual ECG differences to sex/MI status after controlling for anatomical variables. However, regression model errors could distort these estimates. A rigorous evaluation of potential deviations (e.g., variance inflation factors or alternative methods like ridge regression) would strengthen the conclusions.

(3) The manuscript's highly quantitative presentation may hinder readability. Simplifying technical descriptions and improving figure clarity (e.g., separating superimposed bar plots in Figures 2-4) would aid comprehension.

(4) Given established sex differences in QTc intervals, applying the same analytical framework to explore QTc's dependence on sex and anatomy could have provided additional clinically relevant insights.

We thank Reviewer 1 for their kind and constructive comments. While we have thoroughly addressed all specific recommendations below, in brief, we have added new analysis of the variance inflation factor in Supplementary Tables 2 and 3 to reassure readers that the chosen parameter sets exhibit low levels of collinearity, and provided more explanation for why the relative positional parameters were chosen to avoid this issue. We have added explanatory figures for all positional and orientational parameters to improve understanding of the technical details, and improved clarity of existing figures as detailed below. We welcome the suggestion to add QT interval to the manuscript – whilst this was only available in the UK Biobank for a single lead, we have included an analysis of both QT and QTc intervals in this lead to Page 10, and added some discussion of this to the second full paragraph of Page 14.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Comment 1: “Collinearity and Regression Analysis: It would be valuable to assess the collinearity among the regressed parameters (e.g., cardiac size, torso volume, heart center positions [x, y, z], and cardiac orientation angles) and evaluate whether alternative regression methods (e.g., ridge regression) might improve robustness. Additionally, cardiac digital twinning with electrophysiological models could help isolate the exact contribution of electrophysiology while enabling sensitivity analysis. Nonlinear regression or machine learning approaches might also enhance the predictive power of the analysis.”

We thank the reviewer for drawing attention to the important issue of collinearity in the parameter sets used in the regression analysis. To address this, we have added Supplementary Tables 2 and 3, which detail the variance inflation factors for each of the parameter sets used. This was considered in the selection of anatomical parameters – e.g. using relative position not absolute distances between landmarks, which would be more collinear. As these are all below a value of 3.4, we believe that the effect of collinearity is limited, and thus to reduce subjectivity of parameter selection in more complex methods, and encourage interpretability, we have retained our linear regression analysis. In addition, we have added an explanation to the second full paragraph on Page 6 of how we calculated the relative, rather than absolute position of the cardiac centre partially to avoid the problem of collinearity when using multiple absolute distances. We concur that modelling and simulation techniques are well suited to explore the electrophysiological component further – as this is out of the scope of this work, we have addressed the role of these methods in future work in the final paragraph of Page 16.

Comment 2: “Figure Clarity (Bar Plots): The superimposed bar plots in Figures 2-4 are difficult to interpret; separating the bars for each coefficient would improve readability.”

We accept that the stacked bar plots could be improved in their clarity. Whilst plotting each anatomical parameter separately multiplies the number of plots by a factor of nine, and makes comparison between parameters more difficult, we have added clear horizontal grid lines in order to make values easier to read and interpret.

Comment 3: “Feature Extraction Visualization: A schematic figure illustrating the steps for measuring heart positional parameters (e.g., with example annotations) would help readers better understand the feature extraction methodology.”

We agree with the reviewer that the calculation of positional and orientational parameters is crucial to illustrate clearly. We have included additional Supplementary Figures 2 and 3 to better convey these parameters.

Reviewer #2 (Public review):

Summary:

Missed diagnosis of myocardial ischemia (MI) is more common in women, and treatment is typically less aggressive. This diagnosis stems from the fact that women's ECGs commonly exhibit 12 lead ECG biomarkers that are less likely to fall within the traditional diagnostic criteria. Namely, women have shorter QRS durations and lower ST junction and T wave amplitudes, but longer QT intervals, than men. To study the impact, this study aims to quantify sex differences in heart-torso anatomy and ECG biomarkers, as well as their relative associations, in both pre- and post-MI populations. A novel computational pipeline was constructed to generate torso-ventricular geometries from cardiac magnetic resonance imaging. The pipeline was used to build models for 425 post-myocardial infarction subjects and 1051 healthy controls from UK Biobank clinical images to generate the population.

Strengths:

This study has a strength in that it utilizes a large patient population from the UK Biobank (425 postMI and 1051 healthy controls) to analyze sex-based differences. The computational pipeline is stateof-the-art for constructing torso-ventricular geometries from cardiac MR and is clinically viable. It draws on novel machine learning techniques for segmentation, contour extraction, and shape modeling. This pipeline is publicly available and can help in the large-scale generation of anatomies for other studies. This allows computation of various anatomical factors (torso volume, cavity volume, etc), and subsequent regression analysis on how these factors are altered before and after MI from the 12-lead ECG.

Weaknesses:

Major weaknesses stem from the fact that, while electrophysiological factors appear to play a role across many leads, both post-MI and healthy, the electrophysiological factors are not stated or discussed. The computational modeling pipeline is validated for reconstructing torso contours; however, potential registration errors stemming from ventricular-torso construction are not addressed within the context of anatomical factors, such as the tilt and rotation of the heart. This should be discussed as the paper's claims are based on these results. Further analysis and explanation are needed to understand how these sex-specific results impact the ECG-based diagnosis of MI in men and women, as stated as the primary reason for the study at the beginning of the paper. This would provide a broader impact within the clinical community. Claims about demographics do not appear to be supported within the main manuscript but are provided in the supplements. Reformatting the paper's structure is required to efficiently and effectively present and support the findings and outcomes of this work.

We thank Reviewer 2 for their considered and detailed feedback. We greatly appreciate the invitation to elaborate on the electrophysiological factors, and we have added discussion of this matter to the second and third full paragraphs on Page 14, extending to Page 15 and first full paragraph on Page 15, and highlighted the role of modelling and simulation in future work on the third full paragraph of Page 16. We agree that registration errors are one reason behind remaining reconstruction errors and feel a strength of our study is that the large number of subjects used aided in reducing the effect of this noise, and have updated the second full paragraph of Page 16 to reflect this. We are wary of moving too many supplemental figures and tables describing demographic trends to the main manuscript for fear of diluting the specific answers to our research questions. We have however actioned the suggestions as detailed below to reformat the paper, including redressing the balance of supplemental versus main methodological sections, and thank the reviewer for their guidance in increasing our clarity.

Reviewer #2 (Recommendations for the authors):

(1) Please detail what "chosen to be representative of the underlying dataset" means in terms of a validation dataset.

We thank the reviewer for addressing the lack of clarity in this matter. We have added a reference in the third full paragraph on Page 6 to Supplementary Appendix 1.1, where we have included full details of the selection criteria.

(2) “Current guidelines ... further research [16]." The paragraph should begin with a broader statement that is relevant to the fact that the entire body of work focuses on ECG-based diagnosis differences in women, rather than LVEF through echocardiography.

We have revised the introduction to Paragraph 3 on Page 3 to clarify our motivation for focusing on the ECG in order to shape proposals for novel ECG-based risk stratification tools.

(3) The last paragraph of the introduction should more clearly state what was performed and how you aim to prove your hypothesis. There is no mention of the data, the regression model, or other key aspects important to the reader.

We have added methodological details to Paragraph 5 on Page 3 in order to clarify our approach in testing our hypothesis.

(4) An overview paragraph should be included in the Methods at the beginning.

We thank the reviewer for this valuable suggestion – we have added an overview paragraph to the start of the methodology section on Page 5.

(5) The computational pipeline portion of the methods should be written in full paragraphs instead of almost a bulleted list. In general, more details from the supplement should be provided in the methods.

We thank the reviewer for raising important points concerning the balance of methodological description in the main manuscript and the supplementary materials. We have added detailed description of the reconstruction pipeline to Pages 5 and 6. We feel that the ordered format of the methods section adds to the reproducibility and transparency of our methodology.

(6) The torso reconstruction method was already validated in Smith et al. [29]. What value does your additional validation bring to this methodology? Furthermore, how does the construction of the ventricular-torso reconstructions using the cardiac axes (not just the torso contours) influence ECG metrics?

We apologise that this was not clear – we have clarified in Paragraph 4 on Page 5 that while Smith et al. 2022 provided a detailed validation to the contour extraction networks, it did not validate the torso reconstruction pipeline, as it only presents the reconstruction of two cases as a proof of concept. We have also expanded the second full paragraph on Page 6 to explain that the sparse (but not dense) cardiac anatomies were constructed in order to calculate the cardiac size, which we found was a key factor moderating many ECG biomarkers. We also specified that the cardiac position and orientation were necessary in order to relate these to the torso axes and positions of the ECG electrodes.

(7) Include the details of the regression analysis in the main body of the methods for the readers. This is crucial to the claims and outcomes of the paper. Only a sentence is included in the results and one in the figure: "Each factor's contribution is calculated from the product of the regression coefficients and anatomical sex differences (Supplementary Appendix 1.5)." What specific contributions can I expect to see in the results figures? The results are filled with methodological aspects that should be in the results.

We thank the reviewer again for this important comment regarding the balance of the main text methodology and supplementary methodology sections. We have added detail to the statistical analysis section of the main text on Pages 7 and 8 in order for the reader to understand the following results section without consulting the supplemental methods. We have also removed these details from the results section.

(8) What is "the remaining estimated effect of electrophysiology". Did you do simulations on the electrophysiology, or how is this computed from the clinical data of patients? More explanation is needed, as without this, the paper is just focusing on anatomy.

We have clarified this important point by moving the explanation of the methodology underpinning our estimation of the electrophysiological contributions using the clinical ECGs from the supplementary methods to the main manuscript on the second full paragraph on Page 7, and continuing to Page 8. We have also specified the role of simulations studies in future work on the final paragraph on Page 16.

(9) Include an overview paragraph of the methods to create more structure.

We thank the reviewer again for the further attention to this issue – as previously, we have added an overview paragraph to the methodology section on Page 5.

(10) Only 19.8% of the patients were female, which is probably due to females having a more severe presentation of the disease. How does this impact, bias, or skew your results?

This comment raises a very interesting point, and while the origin of this imbalance is of course multifactorial – women likely do have lower rates of MI events due to the cardioprotective role of estrogen and different health promoting behaviours, and our sex imbalance was reflective of wider trends in MI diagnosis. However, as mentioned in Paragraph 2 Page 3 of the text, there are more missed MI diagnoses in women, and we agree that this may lead to a more severe presentation of female MI pathophysiology. We have expanded the first full paragraph on Page 16 to specify the ECG and demographic impacts that this has on our results, and that it is a strength of this work that we may contribute to future adjustment of the diagnostic criteria, such that future investigations do not have this bias, and that clinical outcomes are improved.

(11) A lot of extra information is provided in Tables 1 and 2. Include additional information in the supplements that is not directly relevant to your findings.

We agree that Table 2 is supplementary, rather than critical information, and have moved it accordingly to the Supplementary Materials on Page 38. We do believe that Table 1 is central for understanding the extracted dataset.

(12) Combine paragraphs 3 and 4 into a single paragraph. "Current guidelines..." and "T wave amplitude...". They are part of a single coherent concept.

We have removed the paragraph break on Page 3 Paragraph 3.

(13) Check all acronyms throughout the paper. The abbreviation for sudden cardiac death (SCD) is only used once in the same paragraph. Remove the acronym and type it out. T-wave amplitude (TWA) is introduced twice in a Figure caption and not introduced until the methods.

Many thanks for this suggestion – we have reviewed all acronyms in the manuscript.

(14) "Figure 1B showcases the capability of the computational pipeline to extract torso contours and reconstruct them into 3D meshes". Isn't this Figure 1A?

We apologise that this was unclear, and have updated the sentence on the first full paragraph of Page 8 to clarify the purpose of Figure 1B.

(15) No need to state: "Female y-axis limits have been adjusted by the difference in healthy QRS duration between sexes for ease of comparison" in the Figure 2 caption.

We have removed this statement on all relevant captions.

(16) The paragraph "For lead V6, 15.9% of healthy subjects..." can be combined with the previous section.

We have removed this paragraph break on Page 9 to improve readability.

(17) The only demographics I could find were age and BMI. State which demographics you used explicitly. This is especially true when the discussion makes claims like "Our findings suggest that corrected QRS duration taking into consideration demographics...". How did you take them into account?

We accept that our previous description of the demographic adjustment to QRS duration in the discussion did not adequately reflect the comprehensiveness of our approach, and have adjusted the second paragraph on Page 14 to rectify this.

(18) The results section is also almost a bulleted list that should be written and reformatted into paragraphs.

The ordered style of our results section was designed to compare how our obtained data answers our research question differently for ECG intervals, amplitudes, and axis angles. Whilst we have adjusted paragraph breaks and moved methodological details to more appropriate sections, we have retained this stylistic choice.

(19) The following sentence should be in the introduction: "Alterations to the polarity and amplitude of the T wave are used in the diagnosis of acute MI [42] and TWA affects proposed risk stratification tools, particularly markers of repolarization abnormalities [9, 43]."

We thank the reviewer for this suggestion. We have included the discussion of how TWA is separately used in proposed risk stratification and current diagnostic tools in Paragraph 3 of Page 3.

Version 2

Abstract

The electrocardiogram (ECG) is used for diagnosis and risk stratification in myocardial infarction (MI). Women have a higher incidence of missed MI diagnosis and complications following infarction, and to address this we aim to provide quantitative information on sex-differences in ECG and torso-ventricular anatomical features and their interdependence. A novel computational automated pipeline is presented enabling the three-dimensional reconstruction of torso-ventricular anatomies for 425 post-MI subjects and 1051 healthy controls from UK Biobank clinical images. Regression models were created relating torso-ventricular and ECG parameters. We found that female hearts were positioned more posteriorly and superiorly than male, and in MI hearts were oriented more horizontally, especially for women. Post-MI women exhibited less QRS prolongation, requiring 27% more prolongation than men to exceed 120ms. Only half of the sex difference in QRS duration was associated with smaller female cavities. Lower STj amplitude in women was striking, associated with smaller ventricles, but also more superior and posterior cardiac position. Post-MI, T wave amplitude and R axis deviations were more strongly associated with posterior and horizontal cardiac positioning in women than in men. Our study highlights the need to quantify sex differences in anatomical features, their implications in ECG interpretation, and the application of clinical ECG thresholds in post-MI.

Introduction

The electrocardiogram (ECG) is a vital tool for routine clinical assessment of cardiac electrical abnormalities. Patient differences in anatomy and electrophysiology both determine the ECG. Computational and experimental studies have demonstrated that torso-ventricular orientation can substantially affect the ECG, potentially confounding disease diagnosis, and anatomy rather than electrophysiology is the strongest contributor to ECG variability in healthy subjects [1–4]. Sex and body mass index (BMI) are strongly associated with differences in torso and heart anatomy, and thus it is crucial to take them into consideration in ECG interpretation [5–7]. ECG abnormalities are caused by alterations in the electrophysiological substrate, but importantly they are modulated by anatomical factors such as heart size, location and orientation with respect to the ECG leads [2, 3].

Ischemic heart disease is the leading single cause of mortality worldwide, killing 0.1% of the global population per year [8]. Therefore, effective patient diagnosis of myocardial infarction (MI) and risk stratification for treatment selection are crucial for survival and quality of life [9]. There is a higher incidence of missed diagnosis of acute MI in women, and hospital care can be less aggressive [10–12]. This is a potential contributor to the higher female mortality following MI [13], alongside factors such as differing demographic risk characteristics and comorbidities [14, 15].

Current guidelines for assessing arrhythmic risk post-MI use primarily mechanical markers and invasive testing [16]. They fail to identify the majority of sudden cardiac death cases, prompting calls for research into alternatives [17]. Several ECG-based biomarkers have been proposed building on mechanistic knowledge of the underlying electrophysiological substrate post-MI. QRS duration (QRSd) and its dispersion have been a key component of proposed electrocardiographic risk stratification following MI. However, findings of its effectiveness are mixed, with some trials suggesting it is a predictor of arrhythmic risk following MI [18, 19], while others showing it is more likely a predictor of general mortality than arrhythmic inducibility [20, 21]. T wave amplitude (TWA) and ST segment changes are also important, reflecting repolarisation heterogeneity [22]. Both of these biomarkers are included in current criteria for diagnosis and patient risk stratification in acute MI, with ST-elevation MI (STEMI) mandating immediate reperfusion therapy [23]. More recently, markers of repolarisation heterogeneity have also shown promise in risk stratification following MI [24, 25].

Women’s ECGs typically exhibit shorter QRSd and lower ST junction (STj) and TWA, but longer QT intervals, than male [26]. Both electrophysiological and structural factors contribute to these differences. Testosterone increases repolarising potassium currents and decreases L-type calcium currents, both acting to speed up repolarisation [5]. Additionally, left ventricular (LV) mass is smaller in women than men [27], explaining in part the shorter QRSd and lower STj and TWA [28, 29]. Smaller torso volume can also increase amplitudes of the QRS and T waves in women [1, 2]. However, the impact on STj amplitude and whether anatomy is the only factor explaining these differences is unknown. In women, electrophysiological remodelling post-infarction may be less substantial than in men, partially from the protective effects of estrogen [30]. However, sex differences are often overlooked in the interpretation of ECG biomarkers, and thus investigations are urgently needed to fill these gaps for improvements in patient diagnosis, risk stratification, and treatment selection.

This study aims to quantify sex-differences in heart-torso anatomy and ECG biomarkers, as well as their relative associations, in healthy and post-MI subjects. To enable a population-wide study of heart-torso anatomy, we first develop and validate an automated image processing pipeline to reconstruct three-dimensional torso anatomical models using clinically standard cardiac magnetic resonance (CMR) images. We then apply the methodology to a cohort of 1476 UK Biobank (UKB) participants to statistically relate extracted anatomical parameters to relevant ECG biomarkers, including the effects of sex and MI status differences. We hypothesise that sex-differences in ECG biomarkers are only partially explained by smaller female heart and torso anatomies, and that sex-differences in electrophysiology and torso-heart position and orientation are particularly important, especially post-MI. These insights will lay the groundwork for a process to correct for the impact of torso-ventricular anatomy on the ECG, enabling personalised, automated ECG interpretation.

Methods

Overview

A novel pipeline was first developed and validated to reconstruct torso anatomy from CMR imaging, which enabled the extraction of anatomical biomarkers describing the geometric relationship between the heart and torso. These anatomical biomarkers were then statistically related to relevant clinical ECG biomarkers in both healthy and post-MI populations in order to assess the effect of anatomical sex differences on the ECG.

Computational pipeline

Figure 1A illustrates the fully automated pipeline used to generate personalised torso anatomical reconstructions from clinically standard CMR imaging. This integrates and reformulates state-of-the-art image processing techniques in order to give the first fully automated, and extensively validated, torso-ventricular reconstruction pipeline of its kind.

Structure and validation of novel, automated torso reconstruction pipeline.

A: proposed end-to-end automated 3D torso reconstruction pipeline from 2D standard clinical cardiac magnetic resonance scans. The torso contours are first extracted from the images. A statistical shape model (SSM) is fitted to the contours, which is then optimally deformed. a: spin of the cardiac short axis plane around the torso vertical axis, 0: verticality of the cardiac long axis, γ: tilt of the cardiac short axis plane. B: (i) comparison of Dice coefficient between single stage contouring (blue) and 3 stage segmentation, postprocessing and refinement (red). (ii) surface-to-contour distance for an example case between the automatically reconstructed surface and the automatically generated torso contours. (iii) surface-to-surface distance between the torso mesh created using the automated pipeline and the manually annotated contours for the subjects with the smallest (left) and largest (right) electrode error respectively.

The pipeline combines machine learning-based segmentation and contour extraction, statistical shape modelling and automated mesh deformation in the following steps:

Three-stage torso contour extraction from CMR imaging using machine learning to form sparse torso representation: Approximately 60 localiser and scout images per subject were first segmented using a convolutional neural network, then their outline was extracted with automated post-processing, and finally refined using a second network. U-net [31] was used for both network architectures; detailed description of the architecture of these network, and the validation of the torso extraction (but not reconstruction) have been described in Smith et al. 2022 [32].

Reconstruction: To generate the initial 3D torso mesh, a statistical shape model (SSM) was applied over the automatically extracted sparse torso contours in 3D space, which was then optimally deformed by minimising the distance to the extracted contours to produce the final 3D torso mesh.

Fitting the SSM: In order to generate the torso meshes from sparse torso contours, a highresolution SSM of human body shapes was employed for the initial 3D torso reconstruction. The SSM was generated from 3D optical body surface scanning of 4308 subjects in the CAESAR study, the largest commercially available scan database to date [33]. Detailed information on the SSM generation is provided in Pishchulin et al. 2017 [34]. As the initial condition, the mean SSM was translated into the subject’s coordinate system by positioning the centre of heart from the mean SSM onto the centre of the subject’s heart. The latter centre was approximated by the closest point to the intersection lines between the cine short axis (SAX) and long axis (LAX) slices. Over the automatically extracted sparse torso contours in 3D space, the SSM was fitted by the optimal estimation of principal components, followed by rigid transformation, such that: minq,R∑d(Ci,R(X¯+Φq))2

where Ci is the ith point on sparse torso representation, X¯ the mean shape model, Φ = (ϕ1,ϕ2, ...) the SSM modes of variations, q the estimated set of parameters along the principal modes, and R the rigid transformation. The fitting of the reconstructions was restricted to the first 40 principal components of the SSM.

Deforming for patient-specific torso reconstruction: The initial generated torso mesh is inherently limited by the variability in the SSM. Hence, in order to accurately capture subject-specific variations, a final deformation step was performed via approximate thin plate splines (TPS) [35] to produce the final 3D torso mesh, minimising the distance to the extracted contours. In case the automated torso contour extraction generated sparse outlier points, any torso points over the 95th percentile were removed.

The initial deformation field was computed by considering the closest point on the SSM-generated mesh from the points on the sparse representation. A series of small, smooth deformations using approximate TPS was then applied to push the 3D mesh closer to the torso points, while preserving the smoothness and local topological properties [36]. Let F be the deformation field such that: argminFλ∑i(||F(Pi)−Q′i||)2+J23(F)and Q′i=Pi+β(Q1−Pi)‖Qi−Pi‖

where {Pi} is the set of closest points on the surface Mt to the torso points {Qi} on the sparse model, and J23 is the TPS functional using derivatives of order 2 and image dimensions 3. The deformation Mt+1 = F(Mt) was iteratively applied in a diffeomorphic manner, resulting in a composition of several smooth approximations morphing initial mesh M towards the sparse representation. Laplacian smoothing, decimation, and affine transformation were applied in the end to ensure the local geometric and topological characteristics of the reconstructed mesh. Finally, the torso surface was remeshed with a restricted Frontal-Delaunay algorithm using the mesh generator JIGSAW [37] with a specified element size of 1cm.

Integration with a sparse cardiac mesh and subsequent extraction of torso-ventricular anatomical parameters: Anatomical data for the size, position, and orientation of the heart were obtained in order to describe the geometrical relationship between the heart and torso. Cardiac surfaces were segmented using the method in Banerjee et al. 2021 [38] to give sparse cardiac anatomies, and thus calculate the cardiac size. Torso-ventricular anatomical parameters were extracted - full details for the calculation of each parameter are given in Supplementary Appendix 1.2. The torso volume and two metrics of cardiac size were estimated. The locations of the ECG electrodes were identified on the mean SSM torso, and their positions were transformed with the mesh such that the resulting torso had electrodes located in equivalent locations. As illustrated in Supplementary Figure 2, the position of the heart centre relative to the electrodes was estimated in the x (lateral), y (posterior) and z (superior) directions. Relative (dimensionless) parameters describing the percentage of the cardiac centre position between the most lateral and the most medial electrode were used in order to avoid the problem of collinearity with using multiple absolute distance measures, and similarly for the posterior and superior position. The orientation of the heart was quantified by the Euler angles of the cardiac axes with respect to the torso axes, shown in Supplementary Figure 3. This includes the spin of the cardiac short axis plane around the torso vertical axis, the verticality of the cardiac long axis, and the tilt of the cardiac short axis plane.

Validation of the torso reconstruction pipeline

The CMR images of 30 subjects, chosen to be representative of the underlying dataset (as detailed in Supplementary Appendix 1.1), were manually annotated to form an independent test set for validation. Final results of the three-stage torso contour extraction process were compared with simply using a single u-net architecture to process raw images directly to refined contours in order to assess the accuracy of the torso-air boundary placement. Separate reconstructions of the 30 test subjects were made using the fully automated pipeline and using manually annotated contours. For each subject, the surface-to-surface distance between the two reconstructions was calculated to evaluate error propagation from contouring to reconstruction, as detailed in Supplementary Appendix 1.3. The mean surface-to-contour distance was calculated for each test subject to evaluate the quality of the reconstruction.

Dataset

The UKB dataset [39] was chosen for its high quality CMR imaging and 12-lead ECGs in a large population. 1646 control subjects were randomly selected from the UKB cohort alongside the 479 subjects with a history of MI, as defined in Supplementary Appendix 1.1. The healthy cohort was made by excluding control subjects with disease diagnoses, and subjects in either population were excluded if their ECG or imaging data was not complete, as described in Supplementary Appendix 1.1. The final population for reconstruction and statistical analysis comprised 1051 healthy and 425 post-MI subjects (total 1476), as shown in Table 1 and Supplementary Table 1. 92.7% of subjects identified as White British. The post-MI population was 6.1±0.4 years older and had 2.4±0.2 kg/m2 higher BMI, with no statistically significant differences in ethnicity. This data was obtained through UK Biobank Resource under Application Number ‘40161’ and the data was processed according to their guidelines. All participants gave written informed consent before enrolment in the UKB cohort.

Statistical analysis

ECG parameters were obtained from the UKB database, which were automatically extracted with the proprietary software Cardiosoft [40]. QT and QTc interval were only available for lead I. The distribution of each anatomical and ECG parameter was compared between populations as in Supplementary Appendix 1.4. STj amplitude refers to the amplitude of the ST segment measured at the junction point (the end of the QRS complex). Each ECG parameter was separately linearly regressed against the set of anatomical parameters (chosen to minimise collinearity) for each subpopulation. This includes the cardiac and torso size, the relative position of the cardiac centre with respect to the electrodes in the lateral, posterior and superior directions and the three cardiac orientation parameters (spin, verticality, and tilt), all as previously defined. Additional linear models that also included the categorical sex parameter were obtained for mixed sex populations, and multiplied by the difference in the parameter between men and women to estimate their contribution to the ECG biomarker sex difference. The effect of electrophysiology was estimated as the coefficient of the sex variable after all anatomical parameters had been controlled for. The contributors to the difference between healthy and post-MI populations were similarly calculated, separately for men and women. A linear model was trained to estimate, and therefore correct, the total anatomical contribution to ECG parameters using only demographic features (sex, height, BMI, and age), as described in Supplementary Appendix 1.4.

Results

Reconstructed torso surfaces matched input contours to within 2.19mm:

Figure 1B illustrates the validation of the torso reconstruction pipeline, showcasing its capability to extract torso contours and reconstruct them into 3D meshes. The three-stage segmentation, automated post-processing, and refinement procedure improved the Dice coefficient of torso contour extraction, compared to a single-stage contouring network, across all image views (Figure 1B(i)). The median surface-to-contour distance, showing the quality of the reconstructed torso surface (Figure 1B(ii)), for the test subjects was 2.19mm (inter-quartile range - IQR: 1.982.37mm). The median surface-to-surface distance between torso meshes reconstructed using the manually annotated contours and the machine learning pipeline for the 30 test subjects was 0.82mm (IQR: 0.69-1.10mm). Figure 1B(iii) shows the reconstructed surfaces for the test subjects with the smallest and largest median reconstruction errors. Both reconstructed surfaces were smooth and anatomically realistic, and, across most of the torsos, errors were within the millimetre range.

Torso volume was larger in post-MI than in healthy subjects in women and men, by 3.7±0.9 dm3 (10.2±3.3%) and 3.5±0.5 dm3 (8.2±1.9%) respectively (Supplementary Figure 4). Additionally, for women, in post-MI the heart was positioned more posteriorly by 3.1±0.9mm, than in healthy subjects. The heart was also oriented with its long axis more horizontally by 6.0±1.2° and 4.3±0.5° in post-MI versus healthy women and men, respectively. These changes were related to the increased BMI of post-MI subjects (Supplementary Figure 4K). Age was also considered (Supplementary Figure 4J), but its effects were considerably less substantial.

QRS duration was shorter in women than in men, particularly post-MI, due to smaller hearts and reduced impact of electrophysiological abnormalities:

For both healthy and post-MI subjects, QRSd was shorter in women versus men in all leads (Figures 2A and 2B), with contributions from each anatomical factor displaying sex differences (Figures 2C and 2D). Sex differences in cavity volume explained approximately half of the sex difference in QRSd in both healthy and post-MI subjects (a mean value across all leads of 3.4±1.3ms out of the 6.0±1.5ms mean total in healthy, and 4.5±1.4ms of a total sex difference of 8.3±2.5ms in post-MI). In healthy subjects, the electrophysiological contribution to QRSd sex differences was similar in all precordial leads (1.0±1.0ms to 2.3±1.2ms). In post-MI subjects, it was considerably more variable (0.2±2.8ms in V1 to 6.3±3.0ms in V6). The total estimated impact of anatomy on the QRSd was over 5ms in lead V6 for 12.4% of healthy subjects. After correction by a linear model that was trained only on demographic features (sex, height, BMI, and age), this proportion reduced to 2.9%.

Sex differences in QRS duration in healthy and post-MI subjects.

A, B: mean QRS duration (QRSd) for each ECG lead in healthy and post-MI subjects respectively with women (red circles) and men (cyan triangles). C,D: contribution of anatomical parameters and electrophysiology to sex differences in QRSd for healthy and post-MI subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between male and female populations). E, F: mean QRSd for each ECG lead in men and women respectively for healthy (black squares) and post-MI (red diamonds) subjects. G, H: contribution of anatomical parameters and electrophysiology to differences in QRSd between healthy and post-MI subjects for male and female subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between healthy and post-MI populations).

Post-MI women exhibited less QRS prolongation than men (Figures 2E and 2F), with mean percentage increase in QRSd from healthy to post-MI, across all leads, of 3.4±2.3% and 5.8±1.5% respectively. This is explained by a reduced contribution of electrophysiological abnormalities post-MI in women, as shown in Figures 2G and 2H (smaller grey bars).

The QT interval was 6.3±2.0ms and 1.1±3.9ms longer in healthy and post-MI women than men respectively in lead I (the only lead with available QT interval data). Figure 3A shows the electrophysiological and anatomical contributions to these differences. While electrophysiological differences prolongated the female QT interval in comparison with male (shown as negative gray contributions), this was partially compensated by the smaller female ventricles (resulting in shorter female QT interval). Figure 3B shows the contributions to the 15.2±1.6ms and 11.0±3.3ms longer QTc interval in healthy and post-MI women respectively than in male. Importantly this shows that the anatomical contribution was substantially reduced – for example from 18.9±5.9ms to 2.5±2.1ms in the healthy population, primarily by near elimination of the contribution from the cardiac size. Additionally including heart rate in the QRSd correction model reduced the proportion of cases where corrected anatomical contribution exceeded 5ms from 2.9% to 2.3%.

Sex differences in raw versus corrected QT interval.

Contribution of anatomical parameters and electrophysiology to sex differences in raw QT (A) and corrected QT (B, QTc) intervals for healthy and post-MI subjects in lead I (calculated by multiplying the regression coefficient for each factor by its mean difference between male and female populations).

ST junction amplitude was lower in women than men in both healthy and post-MI subjects, explained by smaller hearts in more superior and posterior position:

STj amplitude was lower in women versus men, particularly in the septal and anterior leads, in healthy and post-MI subjects (Figures 4A and 4B). As for QRSd, this was partly explained by women’s smaller LV mass (by 29.3±5.4% and 27.1±5.7% in healthy and post-MI subjects respectively), but also by a more superior position in women than men for healthy subjects (by 9.6±0.9mm) and more posterior position for post-MI subjects (by 4.3±1.0mm) (Figures 4C and 4D). Both smaller LV mass and electrophysiology contributed to lower female STj amplitude, whereas a smaller female torso volume was generally associated with a higher STj amplitude (yellow and grey versus orange bars in Figures 4C and 4D). This pattern was consistent for differing measurement times of the ST level (Supplementary Figure 8). Baseline STj amplitude in lead V4 was 0.028±0.003mV lower in healthy women than healthy men. Healthy women with BMI>25kgm-2 had a STj amplitude 0.047±0.004mV lower than healthy men with BMI<25kgm-2, associated with their smaller ventricles, but larger torsos.

Sex differences in ECG amplitudes in healthy and post-MI subjects.

A, B: mean STj amplitude (measured at QRS offset) for each ECG lead in healthy and post-MI subjects respectively for women (red circles) and men (cyan triangles). C, D: contribution of anatomical parameters and electrophysiology to sex differences in STj amplitude for healthy and post-MI subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between male and female populations). E, F: mean T wave amplitude (TWA) for each ECG lead in men and women respectively for healthy (black squares) and post-MI (red diamonds) subjects. G, H: contribution of anatomical parameters and electrophysiology to differences in TWA between healthy and post-MI subjects for male and female subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between healthy and post-MI populations).

Lower T wave amplitude following MI was more strongly associated with a more posterior cardiac position and more horizontal orientation in women than men:

Precordial TWA was lower in women than men, by 41.7±20.3% and 31.2±20.0% in healthy and post-MI populations respectively (Supplementary Figure 9). MI resulted in reduced TWA, in all leads in men, whereas only in V4-6 and limb leads in women (Figures 4E and 4F). For men, this was largely associated with electrophysiological factors substantially reducing TWA across leads (Figure 4G). However, for women, the reduced precordial TWA was largely associated with more posterior and horizontal ventricles from their larger BMI (Figure 4H), and electrophysiological factors actually counteracted the TWA post-MI reduction for septal leads (shown as positive contributions).

R axis deviations following MI were associated with more horizontal cardiac orientation in women, but MI-related electrophysiological abnormalities in men:

Figures 5A and 5B break down the anatomical contributions to R and T axis shifts in post-MI compared with healthy subjects. For men, the electrophysiological contribution dominated their 12.0±2.8° left deviated R axis following MI (Figure 5A). However, for women their 5.3±4.0° left deviation in the R axis was mainly associated with a more horizontal cardiac long axis, with electrophysiological factors playing very little role (Figure 5B). This was partly explained by female R axes being more sensitive to changes in verticality - the regression coefficient between R axis and verticality was higher for women than men (Figure 5C). The female T axis was more right deviated following MI than male, largely associated with electrophysiology (Figures 5A and 5B), with less contribution from orientation (Figure 5D).

Sex differences in ECG axis angles in healthy and post-MI subjects

A, B: contribution of anatomical parameters and electrophysiology to differences in R and T axis angles between healthy and post-MI subjects for male and female subjects (calculated by multiplying the regression coefficient for each factor by its mean difference between healthy and post-MI populations). C, D: normalised regression coefficients showing the association between the verticality of the cardiac long axis and the R and T axis angles respectively for the healthy and post-MI populations. Bars shown in red represent regression coefficients significantly different from 0, at a significance level of 0.05. E, F: normalised regression coefficients showing the association between torso volume and the R and T axis angles respectively for the healthy and post-MI populations.

Torso volume significantly affected R axis for both sexes in healthy, but not post-MI subjects (Figure 5E, regression coefficient of R axis against torso volume). For example, in healthy women, every 10dm3 increase in torso volume was associated with a decrease in R axis of 7.2±3.0°. This was related to an increased distance to the left leg electrode (inversely related to the R axis) without as substantial an increase in distance to the left arm electrode (positively related to the R axis). For healthy women, every 10dm3 increase in torso volume was associated with an increased distance to the heart centre of 19.4±0.7mm and 5.9±0.6mm for the left leg and arm electrodes, respectively. However, in post-MI women, the gap between these differences was smaller, with corresponding increased distances of 16.6±1.8mm and 8.3±1.7mm. This association was less strong for the T wave (Figure 5F).

Discussion

This study provides quantification of sex differences in image-based anatomical factors and their implications for ECG biomarkers in healthy and post-MI subjects, enabled by a novel computational clinical image processing pipeline. With this work, we aim to contribute to addressing the higher incidence of missed diagnosis and increased female MI mortality following infarction [12, 13, 41]. Our analysis of a large cohort of healthy and post-MI subjects’ images and ECG data provides quantitative evidence on the importance of considering demographic characteristics and their anatomical differences, particularly sex, for corrections of ECG biomarkers, key to patient risk stratification.

An important contribution is the novel computational pipeline for automated 3D torso reconstruction, which enables high-throughput investigations into the effects of anatomical variability on the study of a wide range of cardiac diseases. The adaptation of the pipeline to standard clinical cardiac imaging allows for the exploitation of large databases, such as the UKB, and sets the groundwork for clinical tools to personalise ECG interpretation considering patient anatomy.

Our analysis provides the following insights into sex-related anatomical differences of the impact of MI on key ECG biomarkers: 1) Half of the shorter QRSd in women was explained by smaller cavity volumes in both healthy and post-MI subjects. 2) Female STj amplitude was lower than male in all precordial leads for healthy and post-MI subjects, associated with a smaller LV mass and more superior and posterior position. 3) In women, anatomical factors had a stronger impact on TWA changes following MI than in men. 4) R axis was more left deviated in MI than healthy populations; for women this was associated with a more horizontal cardiac long axis, but for men with electrophysiology.

Both clinically and moderately prolonged QRSd have been associated with increased risk following MI [18, 19]. Importantly, our analysis reports that mean QRSd would have had to increase by 27% more for women than men to exceed 120ms, and by 68% more to exceed 100ms, both in lead aVL. Sex differences in QRSd are only partly explained by cardiac size, in line with Mincholé et al. 2019 [1], and with the biophysical understanding that larger distances require longer to traverse [42]. However, even after accounting for anatomical factors, there seems to be an electrophysiological component to the increased male QRSd, which supports previous works finding that men had more complex activation pathways, prolonging their QRSd [43]. Our finding showed that taking into account only simple demographic information (age, sex, BMI, and height) substantially reduced the impact of anatomy on QRSd. This suggests that correction methods without complex reconstruction may be sufficient to capture anatomy-ECG relationships in MI diagnosis and risk stratification. Nonetheless, detailed characterisation of cardiac shape utilising 3D reconstruction has been shown to provide additional prognostic value and improve risk stratification following infarction [44, 45]. QRS prolongation in post-MI subjects has been reported in previous studies, particularly cooccurring with ST changes in leads exploring the affected region [46]. It can be explained by slowed conduction velocity in the affected region, involvement of the conduction system such as the bundle branches and Purkinje, and in the chronic case, with dilation and hypertrophy of non-infarcted myocardium [20, 47, 48]. The fact that the pattern across leads of the electrophysiological prolongation of QRSd following MI differed between men and women potentially points to differences in the location or direction of those conduction disturbances. Our finding that sex differences in QTc had much less contribution from cardiac size than in QT suggests that part of the anatomical masking of the longer QT interval in women is because their smaller hearts naturally beat faster, shortening the QT interval. There is some evidence that QRS duration also shortens with faster heart rate, though to a lesser extent, and the heart rate may serve as an easily obtained proxy for cardiac size, so correction by heart rate may play a role in QRSd-based risk stratification [49].

The finding that STj amplitude was substantially lower in women than men in all septal and anterior leads (V1-V4) is also striking. Based on our findings, the dual effects of cardiac and torso volume need to be considered in the interpretation of ST-elevation through the lens of the anatomical context of the patient. As we considered healthy and post-MI subjects (rather than with acute ischemia), our data reflect the normal baseline of STj amplitude, from which any deviations in acute ischemia can be measured. The Fourth Universal Definition of Myocardial Infarction only has sex-specific thresholds in leads V2 and V3 for identifying ST-elevation, a key component in the diagnosis of STEMI [50]. For example, in lead V4, the clinical threshold for both female and male ST-elevation is 0.1mV [50]. Based on our findings, due to their lower baseline values, overweight healthy women would need a 63% larger increase in STj amplitude in this lead to be classified as ST-elevated than normal weight healthy men. Therefore, this work represents a significant step forward in the characterisation of anatomical influences of its baseline level, from which deviations can then be more accurately identified in the context of anatomical variation. It also highlights the important intersectional effects of demographic characteristics such as sex and BMI, suggesting that both of these factors should be accounted for in assessing ST-elevation to improve accuracy of MI diagnosis and classification. We found a substantial electrophysiological component to the sex difference in STj amplitude in both healthy and post-MI subjects. This potentially happens as the longer female action potential duration causes less overlap between activation and early repolarisation in different sections of tissue occurring at the same time [51]. Again, as this occurred in leads without sex specific MI diagnostic guidelines such as V1 and V4, such thresholds should be reconsidered.

Alterations to the polarity and amplitude of the T wave are used in diagnosis of acute MI [50] and TWA affects proposed risk stratification tools, particularly markers of repolarisation abnormalities [9, 52]. Our data reveal that, particularly in women, a substantial proportion of TWA reductions were associated with anatomical differences, such as the decreased verticality of the post-MI heart. This implies that without anatomical corrections, such as adjusting for BMI, the female TWA may be less representative of its underlying electrophysiology, and therefore less representative of any pathology. This increased effect of anatomical parameters in women was partly due to their larger difference in key anatomical factors such as posterior position and long axis orientation from healthy to post-MI subjects. This reflects their larger increase in BMI, associated with a more horizontal cardiac long axis, as described in previous works, suggesting this was related to an upwards shift in the diaphragm [6, 53]. Furthermore, the female TWA was more sensitive to changes in anatomical parameters. This could be explained by sex differences in the directions of the T wave vectors and nonlinearities in the structure-function relationships. Our finding that TWA was affected by cardiac position and orientation is in line with previous simulation studies [2] albeit in a small number of anatomies (N=5) and without electrophysiological variation. Electrophysiologically, TWA variations are caused by differences in the magnitude and timing of repolarisation gradients, and lower TWA in healthy women than men is in part related to their slower repolarisation [54]. TWA drops, and even inverts, in MI due to disruption of the usual direction of the transmural repolarisation gradient from adverse remodelling, alongside tissue death [55]. Our results found a different pattern of electrophysiologically-driven TWA changes following infarction between men and women. This could point to either differences in the nature of the repolarisation remodelling, or how this remodelling interacts with naturally present sex differences in cellular electrophysiology.

Our data revealed that the verticality of the cardiac long axis was particularly significant in determining R axis angle, especially for women. This represents a step forward in understanding the complex relationship between electrical and anatomical axes [6, 56, 57]. Moreover, the leftdeviation of the R axis following infarction may look similar in both sexes; in men this was representative of an electrophysiological change, presumably alterations to the direction of the QRS vector caused by both hypertrophy and areas of electrical inactivity [47]. However, for women this change was almost entirely associated with anatomical differences. This suggests that it is particularly important to correct R axis angles for anatomical factors in women to avoid misattribution of deviations to electrophysiological pathology. The significant relationship between torso volume and R axis angle for healthy subjects indicates that BMI and height may be critical considerations when using ECG based biomarkers that are sensitive to axis angles. The fact that increased torso volume moved the left leg electrode further away from the heart centre than the left arm electrode suggests that this relationship is caused by a stretching of the anatomical plane in which these angles are viewed. Therefore, this represents another area in which a clinical tool that accounts for anatomical variation in interpreting ECG biomarkers would reduce differential accuracy in risk stratification between demographic groups.

The computational pipeline presented here is available to be applied to further studies. Validation was performed by demonstrating that the extracted contours are near the ground truth torso-air boundary on the clinical imaging, as well as to the final reconstructed surfaces (i.e. high Dice coefficient for contour extraction and low surface-to-contour distances). As detailed in Smith et al. 2022 [32], the three-step segmentation, automated post-processing, and refinement process was particularly effective at removing errors inside the torso at locations of sharp gradient between dark and light, for example due to air in the lungs. Whilst reconstructed 3D torso surfaces were largely smooth and realistic, there were small irregularities, generally on the shoulder and waist regions. These were not typically associated with incomplete CMR artifact removal and may instead result from images being taken on different breath-holds or the image sparsity. The low surface-to-surface distance between reconstructions made using the automatically and manually derived contours demonstrates that the small errors in contour extraction did not lead to significant errors in electrode placement on the torso meshes. Surfaces were generally more accurate on the upper left quadrant of the torso, which is where most electrodes are placed. The validity of the torso-ventricular reconstruction pipeline is further supported as known trends in torso and cardiac size, position, and orientation were qualitatively reproduced, and furthermore were quantitively characterised. This includes the more superior female cardiac position [58, 59] and the horizontal shift in the cardiac long axis for subjects with a larger BMI [6, 53]. As this pipeline uses clinically standard CMR imaging, it can be more easily translated to a clinical tool to correct for the effect of anatomical variation on the ECG in a variety of medical contexts.

Due to the demographics of the UKB [39], this study only included subjects aged 45-80 years old; however, the incidence of MI is very low in people younger than this range [13]. The ethnic background of the dataset also had limited diversity; however, this is roughly representative of the UKB population and thus the ethnic composition of the United Kingdom at the time of recruitment [60]. As databases become more diverse, the mediating effect of ethnic background on the cardiac structure-function relationship should be investigated. Whilst the healthy subjects had a roughly equal sex balance, the post-MI subjects had a male-female ratio of approximately 4:1. Whilst this is in line with sex differences in diagnosed MI events [61], the proportion of female MI events may be underrepresented due to more missed diagnoses, particularly for more mild cases [12]. This may have skewed the diagnosed female post-MI ECG phenotype to be more severely affected, and the demographics to be, for example, older. Taking anatomical sex differences into account when diagnosing acute MI may aid in addressing this imbalance in future MI investigations.

The statistical shape model used for initial torso reconstruction did not have separate male and female models, which may have led to an underestimation of the sex differences in the torsos of male and female anatomies. However, the deformation step of the reconstruction was designed to address this issue by modelling personalised torso variations. The exact electrode positions of the obtained ECGs were not available in the UKB dataset, so these were estimated by placing the electrodes on standard positions on the personalised torso. However, this estimation process allows for use of many similar clinically standard datasets, avoiding complex acquisition protocols. Both voluntary and involuntary patient movement impedes precise determination of the cardiac positioning and the reconstruction of a static torso surface, and is one factor explaining remaining reconstruction errors. The large number of subjects in this study likely reduced the impact of this noise on the statistical results.

As the anatomical factors cannot represent a perfectly complete and independent basis set, the electrophysiological contribution can only be estimated. However, as there is more anatomical variation than can be represented using the given factors, such as the smaller scale geometry of the heart, the role of the anatomy versus electrophysiology was likely under- rather than overestimated. Computational modelling and simulation techniques [62, 63] are well suited to further investigate the impact of electrophysiological sex differences on the ECG in future works. Whilst the study benefits from the electrophysiological and anatomical variation that using clinical data facilitates, it is impossible to fully separate the effects of these factors on ECG biomarkers, as characteristics such as sex affect both categories, making them interdependent. This means that this work complements computational studies, which have less interpersonal variability, but isolate the effect of positional and rotational changes. Scar characteristics were not available in the UKB, but the effect of scar geometry variation could be considered in future studies with the development of scar delineation tools. Future work should focus on the characterisation of ECG biomarkers in MI and other disease conditions. The automated torso reconstruction pipeline this work proposes enables the expansion of the number of anatomies such future computational studies could investigate.

Conclusion

To conclude, this work presents a novel automated pipeline for personalised torso reconstruction and demonstrates its power in exploiting large clinical databases by relating torso-ventricular anatomy and ECG biomarkers for 1051 healthy and 425 post-MI subjects. Results show the considerable influence of anatomical factors in demographic differences of ECG biomarkers. Women and individuals with higher BMI may be disproportionately affected. This underlines the importance of, and lays the foundations for, personalised ECG interpretation that considers a subject’s individual torso-ventricular anatomy, facilitating improvement in diagnosis and risk stratification tools.